---
title: "Lab 1: Census Data Quality for Policy Decisions"
subtitle: "Evaluating Data Reliability for Algorithmic Decision-Making"
author: "Arzoo"
date: today
format: 
  html:
    code-fold: false
    toc: true
    toc-location: left
    theme: cosmo
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

# Assignment Overview

## Scenario

You are a data analyst for the **\[Your State\] Department of Human
Services**. The department is considering implementing an algorithmic
system to identify communities that should receive priority for social
service funding and outreach programs. Your supervisor has asked you to
evaluate the quality and reliability of available census data to inform
this decision.

Drawing on our Week 2 discussion of algorithmic bias, you need to assess
not just what the data shows, but how reliable it is and what
communities might be affected by data quality issues.

## Learning Objectives

-   Apply dplyr functions to real census data for policy analysis
-   Evaluate data quality using margins of error
-   Connect technical analysis to algorithmic decision-making
-   Identify potential equity implications of data reliability issues
-   Create professional documentation for policy stakeholders

## Submission Instructions

**Submit by posting your updated portfolio link on Canvas.** Your
assignment should be accessible at `your-portfolio-url/labs/lab_1/`

Make sure to update your `_quarto.yml` navigation to include this
assignment under an "Labs" menu.

# Part 1: Portfolio Integration

Create this assignment in your portfolio repository under an
`labs/lab_1/` folder structure. Update your navigation menu to include:

```         
- text: Assignments
  menu:
    - href: labs/lab_1/your_file_name.qmd
      text: "Lab 1: Census Data Exploration"
```

If there is a special character like a colon, you need use double quote
mark so that the quarto can identify this as text

# Setup

```{r setup}
# Load required packages (hint: you need tidycensus, tidyverse, and knitr)
library(tidyverse)
library(tidycensus)
library(scales)
library(RColorBrewer)
library(knitr)

# Set your Census API key


# Choose your state for analysis - assign it to a variable called my_state
my_state <- "LA"
```

**State Selection:** I have chosen **Louisiana** for this analysis
because:

Louisiana's rural parishes show the highest income estimate uncertainty,
with low-confidence counties concentrated in areas with smaller, more
dispersed populations. These unreliable income estimates pose a direct
risk to algorithmic systems — parishes with high MOE could be
systematically misclassified for funding priority, meaning communities
with genuine need may be overlooked or incorrectly ranked simply due to
data limitations rather than actual economic conditions.

# Part 2: County-Level Resource Assessment

## 2.1 Data Retrieval

**Your Task:** Use `get_acs()` to retrieve county-level data for your
chosen state.

**Requirements:** - Geography: county level - Variables: median
household income (B19013_001) and total population (B01003_001)\
- Year: 2022 - Survey: acs5 - Output format: wide

**Hint:** Remember to give your variables descriptive names using the
`variables = c(name = "code")` syntax.

```{r county-data}
# Write your get_acs() code here

county_level <- get_acs( state = my_state, geography = "county", 
                     variables = c( med_hh_inc = "B19013_001", tot_pop = "B01003_001"), year = 2022, survey = "acs5", output = "wide")


# Clean the county names to remove state name and "County" 
county_level <- county_level %>% 
  mutate(county = str_remove(NAME, ", Louisiana"), 
county = str_remove( county, " Parish") )



# Hint: use mutate() with str_remove()

# Display the first few rows
county_level
```

## 2.2 Data Quality Assessment

**Your Task:** Calculate margin of error percentages and create
reliability categories.

**Requirements:** - Calculate MOE percentage: (margin of error /
estimate) \* 100 - Create reliability categories: - High Confidence: MOE
\< 5% - Moderate Confidence: MOE 5-10%\
- Low Confidence: MOE \> 10% - Create a flag for unreliable estimates
(MOE \> 10%)

**Hint:** Use `mutate()` with `case_when()` for the categories.

```{r income-reliability}
# Calculate MOE percentage and reliability categories using mutate()
county_level <- county_level %>% 
  mutate ( MOE_perc = (med_hh_incM/med_hh_incE)*100) %>% 
  mutate( confidence_level = case_when(
    MOE_perc < 5 ~ "High Confidence", 
    MOE_perc >=5 & MOE_perc <10 ~ "Moderate Confidence", 
    MOE_perc > 10 ~ "Low Confidence"
    
  ))
# Create a summary showing count of counties in each reliability category

county_level %>% 
  count(confidence_level, name = "county") %>% 
  mutate( share = county / sum(county) * 100)

# Hint: use count() and mutate() to add percentages
```

## 2.3 High Uncertainty Counties

**Your Task:** Identify the 5 counties with the highest MOE percentages.

**Requirements:** - Sort by MOE percentage (highest first) - Select the
top 5 counties - Display: county name, median income, margin of error,
MOE percentage, reliability category - Format as a professional table
using `kable()`

**Hint:** Use `arrange()`, `slice()`, and `select()` functions.

```{r high-uncertainty}

library(knitr)
library(kableExtra)
# Create table of top 5 counties by MOE percentage
reliability_table <- county_level %>% 
  arrange( desc(MOE_perc)) %>% 
  slice(1:5) %>% 
  select( County = county, 
          Median_Income= med_hh_incE,
          MOE = med_hh_incM, 
          MOE_perc, 
          Reliability = confidence_level)
reliability_table

# Format as table with kable() - include appropriate column names and caption
reliability <- county_level %>%
  arrange(desc(MOE_perc)) %>%
  slice(1:5) %>%
  select(
    county,
    med_hh_incE,
    med_hh_incM,
    MOE_perc,
    confidence_level
  ) %>%
  kable(
    col.names = c(
      "County",
      "Median Household Income",
      "Margin of Error",
      "MOE (%)",
      "Reliability"
    ),
    digits = 2,
    caption = "Top 5 Counties with the Highest Median Household Income MOE Percentages",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  footnote(
    general = c(
      "Margin of Error (MOE) expressed as a percentage of the estimate.",
      "🟢 High Confidence: MOE < 5%",
      "🟡 Moderate Confidence: MOE 5–10%",
      "🔴 Low Confidence: MOE ≥ 10%",
      "Source: American Community Survey 5-Year Estimates"
    ),
    general_title = "Notes:"
  )
reliability

```

**Data Quality Commentary:**

All five parishes with the highest income uncertainty fall into the Low
Confidence category, with Cameron Parish showing the most extreme MOE at
24.22% — meaning the true median household income could be nearly
\$17,000 above or below the reported estimate of \$69,847. An
algorithmic system relying on this income data would be most likely to
misclassify Cameron, Red River, Madison, East Carroll, and St. James
parishes, all of which are small, rural Louisiana parishes where the ACS
samples too few households to produce stable estimates. The combination
of small population sizes and geographic isolation in these parishes
drives up uncertainty, meaning communities that may have genuine
economic need could be systematically ranked incorrectly by any
automated funding allocation system.

# Part 3: Neighborhood-Level Analysis

## 3.1 Focus Area Selection

**Your Task:** Select 2-3 counties from your reliability analysis for
detailed tract-level study.

**Strategy:** Choose counties that represent different reliability
levels (e.g., 1 high confidence, 1 moderate, 1 low confidence) to
compare how data quality varies.

```{r select-counties}
# Use filter() to select 2-3 counties from your county_reliability data
selected_counties <- county_level %>%
  filter(confidence_level %in% c(
    "High Confidence",
    "Moderate Confidence",
    "Low Confidence"
  )) %>%
  group_by(confidence_level) %>%
  slice_max(MOE_perc, n = 1, with_ties = FALSE) %>%
  ungroup()
selected_counties
# Store the selected counties in a variable called selected_counties

# Display the selected counties with their key characteristics
selected_counties %>% 
    select(
      county,
      med_hh_incE,
      MOE_perc, 
      confidence_level
    ) %>% 
  kable(
    col.names = c(
      "County",
      "Median Household Income",
      "MOE (%)",
      "Reliability"
    ),
    digits = 2,
    caption = "Selected Counties to Run a tract level analysis",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  footnote(
    general = c(
      "Margin of Error (MOE) expressed as a percentage of the estimate.",
      "🟢 High Confidence: MOE < 5%",
      "🟡 Moderate Confidence: MOE 5–10%",
      "🔴 Low Confidence: MOE ≥ 10%",
      "Source: American Community Survey 5-Year Estimates"
    ),
    general_title = "Notes:"
  )

# Show: county name, median income, MOE percentage, reliability category
```

**Comment on the output:**

The three selected parishes — Cameron, Jefferson Davis, and Natchitoches
— represent distinct reliability tiers, allowing for a meaningful
comparison of how data quality varies across Louisiana's rural
landscape. Cameron Parish, with its small and geographically dispersed
coastal population, shows the highest income uncertainty, while
Natchitoches reflects moderate confidence typical of small interior
parishes. This selection ensures our tract-level analysis captures the
full range of data quality challenges the department may encounter when
deploying algorithmic tools statewide.

## 3.2 Tract-Level Demographics

**Your Task:** Get demographic data for census tracts in your selected
counties.

**Requirements:** - Geography: tract level - Variables: white alone
(B03002_003), Black/African American (B03002_004), Hispanic/Latino
(B03002_012), total population (B03002_001) - Use the same state and
year as before - Output format: wide - **Challenge:** You'll need county
codes, not names. Look at the GEOID patterns in your county data for
hints.

```{r tract-demographics}
# Define your race/ethnicity variables with descriptive names
tract_level <- get_acs(
  state = my_state, 
  geography = "tract", 
  county = c("069", "023", "053"),
  variables = c(
    med_hh_inc = "B19013_001", 
    tot_pop = "B03002_001", 
    white = "B03002_003", 
    black = "B03002_004", 
    hisp = "B03002_012"
  ), 
  year = 2022, 
  survey = "acs5", 
  output = "wide"
)

# Add readable tract and county name columns using str_extract() or similar
tract_level_clean <- tract_level %>%
  mutate(
    tract_number = str_extract(NAME, "\\d+\\.?\\d*"),
    county_name = str_extract(NAME, "(?<=; )[^;]+(?= Parish)")
  )
# Use get_acs() to retrieve tract-level data
# Hint: You may need to specify county codes in the county parameter


# Calculate percentage of each group using mutate()
# Create percentages for white, Black, and Hispanic populations

tract_level_clean <- tract_level_clean %>% 
  mutate( perc_white = ((whiteE/ tot_popE)*100), 
          perc_black = ((blackE/tot_popE)*100), 
          perc_hisp = ((hispE/tot_popE)*100))




```

## 3.3 Demographic Analysis

**Your Task:** Analyze the demographic patterns in your selected areas.

```{r demographic-analysis}
# Find the tract with the highest percentage of Hispanic/Latino residents
# Hint: use arrange() and slice() to get the top tract
highest_hisp_tract <- tract_level_clean %>%
  mutate(perc_hisp = (hispE / tot_popE) * 100) %>%
  arrange(desc(perc_hisp)) %>%
  slice(1) %>%
  select(tract_number, county_name, tot_popE, hispE, perc_hisp)
highest_hisp_tract

# Calculate average demographics by county using group_by() and summarize()
# Show: number of tracts, average percentage for each racial/ethnic group
county_summary <- tract_level_clean %>%
  mutate(
    perc_white = (whiteE / tot_popE) * 100,
    perc_black = (blackE / tot_popE) * 100,
    perc_hisp = (hispE / tot_popE) * 100
  ) %>%
  group_by(county_name) %>%
  summarize(
    num_tracts = n(),
    avg_perc_white = mean(perc_white, na.rm = TRUE),
    avg_perc_black = mean(perc_black, na.rm = TRUE),
    avg_perc_hisp = mean(perc_hisp, na.rm = TRUE),
    .groups = "drop"
  )

county_summary
# Create a nicely formatted table of your results using kable()

county_summary %>%
  kable(
    col.names = c(
      "Parish",
      "Number of Tracts",
      "Avg % White",
      "Avg % Black",
      "Avg % Hispanic/Latino"
    ),
    digits = 2,
    caption = "Average Demographics by Parish",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  footnote(
    general = c(
      "Percentages represent the average across all census tracts within each parish.",
      "Source: American Community Survey 5-Year Estimates (2022)"
    ),
    general_title = "Notes:"
  )
```

# Part 4: Comprehensive Data Quality Evaluation

## 4.1 MOE Analysis for Demographic Variables

**Your Task:** Examine margins of error for demographic variables to see
if some communities have less reliable data.

**Requirements:** - Calculate MOE percentages for each demographic
variable - Flag tracts where any demographic variable has MOE \> 15% -
Create summary statistics

```{r demographic-moe}
# Calculate MOE percentages for white, Black, and Hispanic variables
# Hint: use the same formula as before (margin/estimate * 100)
tract_level_clean <- tract_level_clean %>% 
  mutate ( MOE_white = ((whiteM/whiteE)*100), 
           MOE_black = (( blackM/ blackE) *100),
           MOE_hisp = ((hispM/hispE)*100))

# Create a flag for tracts with high MOE on any demographic variable
tract_level_clean <- tract_level_clean %>%
  mutate(
    high_MOE_flag = ifelse(
      replace_na(MOE_white > 15, FALSE) |
      replace_na(MOE_black > 15, FALSE) |
      replace_na(MOE_hisp > 15, FALSE),
      "🔴 High MOE (>15%)",
      "Acceptable MOE"
    )
  )
# Use logical operators (| for OR) in an ifelse() statement

# Create summary statistics showing how many tracts have data quality issues
tract_level_clean %>%
  group_by(county_name, high_MOE_flag) %>%
  summarize(num_tracts = n(), .groups = "drop") %>%
  kable(
    col.names = c("Parish", "Data Quality", "Number of Tracts"),
    digits = 2,
    caption = "Data Quality Summary by Parish",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  footnote(
    general = c(
      "A tract is flagged if any demographic variable (White, Black, or Hispanic) has MOE > 15%.",
      "🔴 High MOE (>15%): Estimates may be unreliable.",
      "Acceptable MOE: Estimates are sufficiently reliable for analysis.",
      "Source: American Community Survey 5-Year Estimates (2022)"
    ),
    general_title = "Notes:"
  )
```

## 4.2 Pattern Analysis

**Your Task:** Investigate whether data quality problems are randomly
distributed or concentrated in certain types of communities.

```{r pattern-analysis}
# Group tracts by whether they have high MOE issues
moe_comparison <- tract_level_clean %>%
  group_by(county_name) %>%
  summarize(
    num_tracts          = n(),
    avg_population      = mean(tot_popE,  na.rm = TRUE),
    avg_perc_white       = mean(perc_white, na.rm = TRUE),
    avg_perc_black       = mean(perc_black, na.rm = TRUE),
    avg_perc_hisp        = mean(perc_hisp,  na.rm = TRUE),
    perc_unreliable_white = mean(replace_na(MOE_white > 15, FALSE)) * 100,
    perc_unreliable_black = mean(replace_na(MOE_black > 15, FALSE)) * 100,
    perc_unreliable_hisp  = mean(replace_na(MOE_hisp  > 15, FALSE)) * 100,
    .groups = "drop"
  )
# Calculate average characteristics for each group:
# - population size, demographic percentages

# Use group_by() and summarize() to create this comparison



# Create a professional table showing the patterns

moe_comparison %>%
  kable(
    col.names = c(
      "Parish",
      "Tracts",
      "Avg Population",
      "Avg % White",
      "Avg % Black",
      "Avg % Hispanic",
      "% Tracts Unreliable (White)",
      "% Tracts Unreliable (Black)",
      "% Tracts Unreliable (Hispanic)"
    ),
    digits = 2,
    caption = "Demographic Estimates and Data Reliability by Parish",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  footnote(
    general = c(
      "% Tracts Unreliable = share of tracts where MOE exceeds 15% of the estimate.",
      "High unreliability for minority groups reflects small subgroup populations in these parishes.",
      "🔴 Estimates with high MOE should be interpreted with caution.",
      "Source: American Community Survey 5-Year Estimates (2022)"
    ),
    general_title = "Notes:"
  )
```

**Pattern Analysis:**

The analysis reveals that data quality problems are not randomly
distributed — they are systematically concentrated among minority
demographic groups across all three parishes. Black and Hispanic
populations show substantially higher MOE percentages than white
populations, not because of data collection failures, but because these
groups represent smaller shares of the total population in these
parishes, resulting in smaller ACS sample sizes and therefore greater
statistical uncertainty. This pattern is consistent across Cameron,
Jefferson Davis, and Natchitoches parishes, suggesting a structural
limitation of the ACS in rural, majority-white areas. The practical
implication is serious: an algorithm relying on demographic estimates
for these communities would have the least reliable data precisely for
the minority groups it may be intended to serve.

# Part 5: Policy Recommendations

## 5.1 Analysis Integration and Professional Summary

**Your Task:** Write an executive summary that integrates findings from
all four analyses.

**Executive Summary Requirements:** 1. **Overall Pattern
Identification**: What are the systematic patterns across all your
analyses? 2. **Equity Assessment**: Which communities face the greatest
risk of algorithmic bias based on your findings? 3. **Root Cause
Analysis**: What underlying factors drive both data quality issues and
bias risk? 4. **Strategic Recommendations**: What should the Department
implement to address these systematic issues?

**Executive Summary:**

**Overall Pattern Identification:** Across all four analyses, a
consistent pattern emerges — data reliability in Louisiana's rural
parishes is inversely related to population size and demographic
diversity. County-level income estimates show high uncertainty in
sparsely populated coastal and inland parishes, while tract-level
demographic data reveals that minority group estimates carry
disproportionately high margins of error even within parishes that
appear reliable at the aggregate level.

**Equity Assessment:** Black and Hispanic residents across Cameron,
Jefferson Davis, and Natchitoches parishes face the greatest risk of
algorithmic harm. Because ACS estimates for these groups carry MOE
values well above 15% in the majority of tracts, any algorithm using
these figures to allocate social services would be operating on
fundamentally unreliable inputs for the very communities most likely to
need those services. This creates a feedback loop where data scarcity
compounds existing inequity.

**Root Cause Analysis:** The underlying driver of both data quality
issues and bias risk is the same: small subgroup population sizes in
rural parishes lead to small ACS sample sizes, which in turn produce
wide confidence intervals. This is not a data entry or methodology
failure — it is a structural limitation of survey-based estimation in
low-density, low-diversity areas. Rural minority communities are, by
definition, underrepresented in the sample, making their estimates the
least stable.

**Strategic Recommendations:** The Department of Human Services should
implement a tiered approach to algorithmic decision-making that
explicitly accounts for data reliability. High-confidence parishes can
proceed with automated prioritization, while moderate-confidence
parishes should require human review of algorithmic outputs before
funding decisions are finalized. In low-confidence parishes, and for any
tract where minority demographic estimates carry MOE above 15%, the
department should suspend automated classification entirely and instead
conduct targeted supplemental surveys or partner with local
organizations to gather ground-truth data before making resource
allocation decisions.

## 6.3 Specific Recommendations

**Your Task:** Create a decision framework for algorithm implementation.

```{r recommendations-data}
# Create a summary table using your county reliability data
# Include: county name, median income, MOE percentage, reliability category

# Add a new column with algorithm recommendations using case_when():
# - High Confidence: "Safe for algorithmic decisions"
# - Moderate Confidence: "Use with caution - monitor outcomes"  
# - Low Confidence: "Requires manual review or additional data"

# Format as a professional table with kable()

county_level %>%
  select(county, med_hh_incE, med_hh_incM, MOE_perc, confidence_level) %>%
  mutate(
    recommendation = case_when(
      confidence_level == "High Confidence"     ~ "🟢 Safe for algorithmic decisions",
      confidence_level == "Moderate Confidence" ~ "🟡 Use with caution - monitor outcomes",
      confidence_level == "Low Confidence"      ~ "🔴 Requires manual review or additional data"
    )
  ) %>%
  arrange(desc(MOE_perc)) %>%
  kable(
    col.names = c(
      "Parish",
      "Median Household Income",
      "Margin of Error",
      "MOE (%)",
      "Reliability",
      "Recommendation"
    ),
    digits = 2,
    caption = "Algorithm Implementation Decision Framework by Parish",
    format.args = list(big.mark = ",")
  ) %>%
  kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  ) %>%
  footnote(
    general = c(
      "Recommendations based on Margin of Error (MOE) as a percentage of the estimate.",
      "🟢 High Confidence (MOE < 5%): Estimates are reliable for automated decisions.",
      "🟡 Moderate Confidence (MOE 5–10%): Use with human oversight and outcome monitoring.",
      "🔴 Low Confidence (MOE > 10%): Data unreliability poses risk of algorithmic bias.",
      "Source: American Community Survey 5-Year Estimates (2022)"
    ),
    general_title = "Notes:"
  )
```

**Key Recommendations:**

**Your Task:** Use your analysis results to provide specific guidance to
the department.

-   **Counties suitable for immediate algorithmic implementation:**
    Parishes classified as High Confidence — where median household
    income MOE falls below 5% — are appropriate candidates for
    algorithmic decision-making. In these areas, the ACS estimates are
    stable enough that automated prioritization is unlikely to produce
    systematic misclassification. The department should still monitor
    outcomes over time to detect any emergent disparities.

-   **Counties requiring additional oversight:** Parishes in the
    Moderate Confidence tier (MOE between 5–10%) can be included in
    algorithmic workflows, but only with active human oversight. Program
    officers should review algorithmic outputs for these parishes before
    finalizing funding decisions, and outcome data should be tracked
    quarterly to identify whether any communities are being consistently
    under- or over-served relative to their actual need.

-   **Counties needing alternative approaches:** Low Confidence parishes
    (MOE above 10%), and any tract where minority demographic estimates
    exceed 15% MOE, should be removed from automated decision pipelines
    entirely. For these areas, the department should pursue one or more
    of the following: targeted supplemental ACS surveys, partnerships
    with local nonprofits and service providers to gather administrative
    data, or manual needs assessments conducted by field staff with
    community knowledge.

    1.  Do data reliability patterns persist over time — that is, have
        the same parishes consistently shown high MOE across multiple
        ACS waves (2015, 2017, 2019, 2022) — or do reliability issues
        fluctuate with population shifts and post-disaster displacement
        (particularly relevant for Cameron Parish post-Hurricane Laura)?
    2.  How do tract-level income MOE patterns compare to demographic
        MOE patterns — are the tracts with the least reliable income
        estimates also the tracts with the least reliable racial
        composition estimates, suggesting compounded data quality risk
        for the most vulnerable communities?
    3.  Would switching from ACS 5-year estimates to administrative data
        sources (SNAP enrollment, Medicaid claims, school lunch
        eligibility) produce more reliable proxies for economic need in
        low-confidence parishes, and how would that change which
        communities are prioritized for services?

# Technical Notes

**Data Sources:** - U.S. Census Bureau, American Community Survey
2018-2022 5-Year Estimates - Retrieved via tidycensus R package on
\[date\]

**Reproducibility:** - All analysis conducted in R version \[your
version\] - Census API key required for replication - Complete code and
documentation available at: \[your portfolio URL\]

**Methodology Notes:** \[Describe any decisions you made about data
processing, county selection, or analytical choices that might affect
reproducibility\]

**Limitations:** \[Note any limitations in your analysis - sample size
issues, geographic scope, temporal factors, etc.\]

------------------------------------------------------------------------

## Submission Checklist

Before submitting your portfolio link on Canvas:

-   [ ] All code chunks run without errors
-   [ ] All "\[Fill this in\]" prompts have been completed
-   [ ] Tables are properly formatted and readable
-   [ ] Executive summary addresses all four required components
-   [ ] Portfolio navigation includes this assignment
-   [ ] Census API key is properly set
-   [ ] Document renders correctly to HTML

**Remember:** Submit your portfolio URL on Canvas, not the file itself.
Your assignment should be accessible at
`your-portfolio-url/labs/lab_1/your_file_name.html`
