---
title: "Predicting High-Risk Buildings in Philadelphia"
subtitle: "From Reactive Inspections to Risk-Based Prioritization"
author: "Liz Crouse, Shaurya Chauhan, Arzoo Gupta"
date: "April 28, 2026"
format:
  revealjs:
    theme: serif
    slide-number: true
    progress: true
    transition: fade
    transition-speed: fast
    fig-align: center
    fig-width: 9
    fig-height: 6
execute:
  echo: false
  warning: false
  message: false
editor: visual
---

```{r setup}

library(tidyverse)
library(sf)
library(tigris)
library(tidycensus)
library(lubridate)
library(here)
library(caret)
library(pROC)
library(spdep)
library(knitr)
library(kableExtra)

plotTheme <- theme_minimal() +
  theme(
    plot.title       = element_text(size = 14, face = "bold"),
    plot.subtitle    = element_text(size = 10),
    plot.caption     = element_text(size = 8),
    axis.text.x      = element_text(size = 10, angle = 45, hjust = 1),
    axis.text.y      = element_text(size = 10),
    axis.title       = element_text(size = 11, face = "bold"),
    panel.grid.major = element_line(colour = "#D0D0D0", linewidth = 0.2),
    panel.grid.minor = element_blank(),
    axis.ticks       = element_blank(),
    legend.position  = "right"
  )

mapTheme <- theme_void() +
  theme(
    plot.title        = element_text(size = 14, face = "bold"),
    plot.subtitle     = element_text(size = 10),
    plot.caption      = element_text(size = 8),
    legend.position   = "right",
    plot.margin       = margin(1, 1, 1, 1, "cm"),
    legend.key.height = unit(1, "cm"),
    legend.key.width  = unit(0.2, "cm")
  )
```

```{r load-data}
#| output: false
#| warning: false
#| message: false
inspections <- read.csv("data/BUILDING_CERTS.csv")

block_groups <- block_groups(state = "PA", county = "Philadelphia") %>%
  st_transform(crs = 4326)

inspections_fire_opa <- st_read("data/fire_insp_opa.geojson") %>%
  st_transform(crs = 4326)

phl_311 <- read.csv("data/311reqs.csv")

phl_census <- get_acs(
  geography = "block group",
  variables = c(
    "B19013_001", "B25002_002", "B25002_003", "B25002_001",
    "B25003_001", "B25003_002", "B25003_003", "B01003_001", "B25071_001"
  ),
  state = "PA", county = "Philadelphia", year = 2022,
  geometry = TRUE, output = "wide"
) %>%
  rename(
    Med_Inc         = B19013_001E,
    Occupied_Units  = B25002_002E,
    Vacant_Units    = B25002_003E,
    Total_Units     = B25002_001E,
    Total_Occupied  = B25003_001E,
    Owner_Occupied  = B25003_002E,
    Renter_Occupied = B25003_003E,
    Total_Pop       = B01003_001E,
    Rent_Burden     = B25071_001E
  ) %>%
  mutate(
    Vacancy_Rate        = (Vacant_Units / Total_Units) * 100,
    Pct_Renter_Occupied = (Renter_Occupied / Total_Occupied) * 100,
    Pct_Owner_Occupied  = (Owner_Occupied / Total_Occupied) * 100
  ) %>%
  select(GEOID, NAME, geometry, Med_Inc, Total_Pop,
         Vacancy_Rate, Vacant_Units, Total_Units,
         Pct_Renter_Occupied, Pct_Owner_Occupied, Rent_Burden) %>%
  st_transform(crs = 4326)

inspections <- inspections %>%
  mutate(
    inspectiondate = as.Date(inspectiondate),
    year    = year(inspectiondate),
    month   = month(inspectiondate, label = TRUE),
    quarter = quarter(inspectiondate)
  )
```

```{r build-features}
#| output: false
#| warning: false
#| message: false

opa <- st_read("data/opa_properties_public.geojson") %>%
  filter(sale_price > 10000) %>%
  mutate(year_built = as.numeric(year_built), age = 2026 - year_built)

block_groups_proj <- st_transform(block_groups, st_crs(opa))

opa_sf_bg <- st_join(
  opa %>% filter(category_code_description == "SINGLE FAMILY"),
  block_groups_proj %>% select(GEOID), join = st_within
) %>% st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarize(
    sf_n_properties     = n(),
    sf_avg_market_value = mean(market_value, na.rm = TRUE),
    sf_avg_sale_price   = mean(sale_price,   na.rm = TRUE),
    sf_avg_total_area   = mean(total_area,   na.rm = TRUE),
    sf_avg_age          = mean(age,           na.rm = TRUE)
  )

opa_mf_bg <- st_join(
  opa %>% filter(category_code_description == "MULTI FAMILY"),
  block_groups_proj %>% select(GEOID), join = st_within
) %>% st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarize(
    mf_n_properties     = n(),
    mf_avg_market_value = mean(market_value, na.rm = TRUE),
    mf_avg_sale_price   = mean(sale_price,   na.rm = TRUE),
    mf_avg_total_area   = mean(total_area,   na.rm = TRUE),
    mf_avg_age          = mean(age,           na.rm = TRUE)
  )

opa_comm_bg <- st_join(
  opa %>% filter(category_code_description %in% c("COMMERCIAL", "MIXED USE")),
  block_groups_proj %>% select(GEOID), join = st_within
) %>% st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarize(
    comm_n_properties     = n(),
    comm_avg_market_value = mean(market_value, na.rm = TRUE),
    comm_avg_sale_price   = mean(sale_price,   na.rm = TRUE),
    comm_avg_total_area   = mean(total_area,   na.rm = TRUE),
    comm_avg_age          = mean(age,           na.rm = TRUE)
  )

phl_311_sf <- phl_311 %>%
  filter(agency_responsible == "Fire Department") %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

bg_311 <- st_join(phl_311_sf, block_groups %>% select(GEOID), join = st_within) %>%
  st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarize(n_requests = n()) %>%
  mutate(n_requests = replace_na(n_requests, 0))

bg_insp_summ <- st_join(
  inspections_fire_opa,
  block_groups %>% select(GEOID), join = st_within
) %>% st_drop_geometry() %>%
  group_by(GEOID) %>%
  summarize(
    avg_dst_nr_  = mean(dst_nr_,  na.rm = TRUE),
    avg_dst_cntr = mean(dst_cnt,  na.rm = TRUE),
    avg_dst_trn  = mean(dst_trn,  na.rm = TRUE)
  )

model_data <- inspections_fire_opa %>%
  st_transform(crs = st_crs(block_groups)) %>%
  st_join(block_groups %>% select(GEOID), join = st_within) %>%
  st_drop_geometry() %>%
  left_join(opa_sf_bg,    by = "GEOID") %>%
  left_join(opa_mf_bg,    by = "GEOID") %>%
  left_join(opa_comm_bg,  by = "GEOID") %>%
  left_join(bg_311,       by = "GEOID") %>%
  left_join(bg_insp_summ, by = "GEOID") %>%
  left_join(phl_census %>% st_drop_geometry(), by = "GEOID") %>%
  mutate(
    high_risk         = as.integer(fail == 1),
    n_requests        = replace_na(n_requests, 0),
    sf_n_properties   = replace_na(sf_n_properties, 0),
    mf_n_properties   = replace_na(mf_n_properties, 0),
    comm_n_properties = replace_na(comm_n_properties, 0),
    log_n_comm_properties   = log1p(comm_n_properties),
    log_comm_avg_mv         = log1p(comm_avg_market_value),
    log_comm_avg_age        = log1p(comm_avg_age),
    log_n_sf_properties     = log1p(sf_n_properties),
    log_sf_avg_mv           = log1p(sf_avg_market_value),
    log_sf_avg_age          = log1p(sf_avg_age),
    log_mf_avg_mv           = log1p(mf_avg_market_value),
    log_mf_avg_age          = log1p(mf_avg_age),
    log_avg_vac_units       = log1p(Vacant_Units),
    log_n_requests          = log1p(n_requests)
  ) %>%
  filter(!is.na(high_risk))

set.seed(42)
trainIndex <- createDataPartition(y = model_data$high_risk, p = 0.70, list = FALSE)
train_data <- model_data[trainIndex, ]
test_data  <- model_data[-trainIndex, ]

train_data_balanced <- upSample(
  x = train_data %>% select(-fail),
  y = factor(train_data$fail),
  yname = "fail"
) %>% mutate(fail = as.integer(as.character(fail)))

logit_model_final <- glm(
  fail ~
    avg_dst_nr_ + avg_dst_trn + avg_dst_cntr +
    Vacancy_Rate + Med_Inc + Pct_Renter_Occupied + Rent_Burden +
    log_comm_avg_mv + log_comm_avg_age + log_n_comm_properties +
    log_sf_avg_mv + log_mf_avg_mv + log_mf_avg_age +
    log_n_requests + bldngcrtt +
    avg_dst_nr_:Vacancy_Rate +
    Pct_Renter_Occupied:log_n_requests +
    log_comm_avg_age:Med_Inc +
    avg_dst_cntr:log_comm_avg_mv,
  data   = train_data_balanced,
  family = "binomial"
)

test_data$predicted_prob <- predict(logit_model_final, newdata = test_data, type = "response")
roc_obj <- roc(test_data$fail, test_data$predicted_prob)

threshold_results <- map_df(seq(0.01, 0.99, by = 0.01), function(t) {
  pred_class <- as.integer(test_data$predicted_prob > t)
  if (length(unique(pred_class)) < 2) return(NULL)
  cm <- confusionMatrix(factor(pred_class), factor(test_data$fail), positive = "1")
  tibble(
    threshold   = t,
    sensitivity = cm$byClass["Sensitivity"],
    specificity = cm$byClass["Specificity"],
    accuracy    = cm$overall["Accuracy"],
    kappa       = cm$overall["Kappa"]
  )
})

best_threshold <- threshold_results %>%
  filter(sensitivity >= 0.80) %>%
  arrange(desc(specificity)) %>%
  slice(1)

test_data$predicted_class <- as.integer(test_data$predicted_prob > best_threshold$threshold)
final_cm <- confusionMatrix(factor(test_data$predicted_class), factor(test_data$fail), positive = "1")
```

------------------------------------------------------------------------

## The Problem: A Reactive Inspection System

Philadelphia's building inspection system is driven by **complaints**, not risk:

-   Only **\~7%** of rental units inspected annually *(Pew Charitable Trusts, 2021)*
-   Fewer than **45 inspectors** citywide *(Pew, 2021)*
-   Complaint-driven system → **systematic underreporting** in vulnerable communities *(Urban Institute, 2022)*
-   All buildings treated **equally** despite highly **unequal risk**

> The city lacks a systematic way to prioritize which buildings need inspection most.

------------------------------------------------------------------------

## Why This Matters: Inspection Results {.smaller}

```{r result-distribution}
#| fig-align: center
#| out-width: 100%
#| fig-height: 6.5

inspections_fire_opa %>%
  filter(!is.na(inspctnr)) %>%
  count(inspctnr) %>%
  mutate(
    pct = n / sum(n),
    inspctnr= reorder(inspctnr, n)
  ) %>%
  ggplot(aes(x = inspctnr, y = n, fill =inspctnr)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = scales::percent(pct, accuracy = 0.1)),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_viridis_d(option = "plasma") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15)),
                     labels = scales::comma) +
  labs(
    title    = "Distribution of Building Inspection Results",
    subtitle = "Philadelphia L&I Inspections",
    x = NULL, y = "Number of Inspections",
    caption  = "Source: OpenDataPhilly"
  ) +
  plotTheme
```

These results show that only a small share of inspections fail (\~5%).

However, these failures are not random—they cluster and follow patterns.

→ Opportunity to prioritize inspections based on risk.

------------------------------------------------------------------------

## Failure Rates Vary by Certification Type

```{r cert-type-results}
#| fig-align: center
#| out-width: 85%

inspections_fire_opa %>%
  filter(!is.na(inspctnr), !is.na(bldngcrtt)) %>%
  count(bldngcrtt, inspctnr) %>%
  group_by(bldngcrtt) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = reorder(bldngcrtt, pct), y = pct, fill = inspctnr)) +
  geom_col(position = "stack") +
  coord_flip()  +
  labs(
    title    = "Inspection Results by Certification Type",
    subtitle = "Failure rates vary significantly by certification type",
    x = NULL, y = "Share of Inspections",
    caption  = "Source: OpenDataPhilly"
  ) +
  plotTheme
```

Certification type is an important control—we account for this in the model.

------------------------------------------------------------------------

## The Data: Five Sources Joined at the Block Group {.scrollable .smaller}

| Source | What it contributes |
|------------------------------------|------------------------------------|
| **L&I Building Certifications** | Outcome variable: pass / fail; certification type fixed effects |
| **OPA Property Data** | Market value, age, property count — split by SF / MF / commercial land use |
| **311 Fire Complaints** | Fire department request volume per block group — early warning signal |
| **Census ACS 2022** | Median income, vacancy rate, % renter-occupied, rent burden |
| **Spatial distances** | Distance to nearest inspection (`avg_dst_nr_`), transit, city center |

Each **property-level** observation is enriched by its surrounding **block group context**.

------------------------------------------------------------------------

## Where Do Failures Cluster?

```{r failure-map}
#| fig-align: center
#| out-width: 85%

ggplot() +
  geom_sf(data = block_groups, fill = "grey90", color = "white", linewidth = 0.1) +
  geom_sf(
    data  = inspections_fire_opa %>% filter(fail == 0),
    color = "#2166ac", size = 0.2, alpha = 0.2
  ) +
  geom_sf(
    data  = inspections_fire_opa %>% filter(fail == 1),
    color = "#d6604d", size = 0.5, alpha = 0.6
  ) +
  labs(
    title    = "Spatial Distribution of Inspection Failures",
    subtitle = "Red = Failed (Deficient/Unsafe) · Blue = Passed",
    caption  = "Source: Philadelphia L&I; OpenDataPhilly"
  ) +
  mapTheme
```

------------------------------------------------------------------------

## Modeling Approach: Iterative Logistic Regression {.scrollable .smaller}

**Why logistic regression?** Binary outcome (fail/pass), interpretable odds ratios, handles continuous + categorical predictors and interaction terms.

| Model | Key change | Issue found |
|------------------------|------------------------|------------------------|
| Model 1 | Distance features only, imbalanced data | Near-zero sensitivity — almost no failures predicted |
| Model 2 | \+ ACS features + interactions | Still imbalanced |
| Model 3 | \+ `upSample()` + threshold lowered to 0.3 | Sensitivity 94% but specificity only 35% |
| Model 4 | \+ `bldngcrtt` fixed effects, threshold scan (sensitivity ≥ 0.80) | Near-final |
| **Final** | Dropped noise variables, kept 4 key interactions | **AUC = 0.802** |

------------------------------------------------------------------------

## Final Model Predictors and Key Interactions {.smaller}

::::: columns
::: {.column width="50%"}
**What we capture**

-   **Spatial risk**
    -   Proximity to other inspections\
    -   Distance to transit and city center
-   **Neighborhood context**
    -   Vacancy rate\
    -   Income and rent burden\
    -   Share of renter-occupied housing
:::

::: {.column width="50%"}
**Behavioral & environmental signals**

-   311 complaint activity\
-   Property characteristics (age, value, density)\
-   Certification type (controls for inspection differences)

**Key idea**

-   Risk is shaped by how these factors interact — not any single variable
:::
:::::

------------------------------------------------------------------------

## What Drives Failure? Odds Ratios

```{r odds-ratios-plot}
#| fig-align: center
#| out-width: 85%

or_data <- tribble(
  ~predictor,                            ~OR,    ~direction,
  "Sprinkler cert. (vs baseline)",       3.66,  "Increases risk",
  "Commercial market value",             1.32,  "Increases risk",
  "Multifamily market value",            1.31,  "Increases risk",
  "Multifamily building age",            1.22,  "Increases risk",
  "Vacancy rate (per pct pt)",           1.02,  "Increases risk",
  "Rent burden",                         1.008, "Increases risk",
  "Renter occ. × 311 calls (interact.)", 1.011, "Increases risk",
  "Single-family market value",          0.665, "Decreases risk",
  "Commercial building age (standalone)",0.695, "Decreases risk",
  "311 call volume (standalone)",        0.733, "Decreases risk",
  "Commercial property density",         0.849, "Decreases risk",
  "Median income",                       0.9999,"Decreases risk"
)

ggplot(or_data, aes(x = reorder(predictor, OR), y = OR, fill = direction)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "grey40", linewidth = 0.8) +
  coord_flip() +
  scale_fill_manual(
    values = c("Increases risk" = "#d62728", "Decreases risk" = "#2166ac"),
    name   = NULL
  ) +
  labs(
    title    = "Odds Ratios: Final Logistic Regression Model",
    subtitle = "OR > 1 increases failure risk · OR < 1 decreases failure risk · Dashed = OR of 1",
    x = NULL, y = "Odds Ratio",
    caption  = "Source: Philadelphia L&I; OPA; ACS 2022"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", panel.grid.minor = element_blank())
```

------------------------------------------------------------------------

## Handling Class Imbalance and Threshold Selection {.smaller}

```{r threshold-plot}
#| fig-align: center
#| out-width: 75%

threshold_results %>%
  pivot_longer(cols = c(sensitivity, specificity, kappa),
               names_to = "metric", values_to = "value") %>%
  ggplot(aes(x = threshold, y = value, color = metric)) +
  geom_line(linewidth = 1.2) +
  geom_vline(xintercept = best_threshold$threshold,
             linetype = "dashed", color = "grey40") +
  annotate("text",
           x     = best_threshold$threshold + 0.03,
           y     = 0.45,
           label = paste0("selected t = ", best_threshold$threshold),
           size  = 3.5, color = "grey30") +
  scale_color_viridis_d(option = "plasma", name = "Metric") +
  labs(
    title    = "Performance Across Classification Thresholds",
    subtitle = "upSample() balanced the training set · threshold scanned 0.01–0.99 for sensitivity ≥ 0.80",
    x = "Threshold", y = "Value",
    caption  = "Source: Philadelphia L&I"
  ) +
  plotTheme
```

::: smaller
We selected a threshold that prioritizes identifying unsafe buildings.

This ensures we capture most high-risk properties, even if it means inspecting some low-risk ones.
:::

------------------------------------------------------------------------

## Final Model Performance

```{r final-metrics-table}
tibble(
  Metric = c("AUC", "Sensitivity", "Specificity", "NPV", "Threshold"),
  Value  = c(
    round(auc(roc_obj), 3),
    round(final_cm$byClass["Sensitivity"], 3),
    round(final_cm$byClass["Specificity"], 3),
    round(final_cm$byClass["Neg Pred Value"], 3),
    best_threshold$threshold
  ),
  Interpretation = c(
    "Good discrimination — model meaningfully separates high and low risk properties",
    "Catches 81% of actual failures — meets our target for a triage tool",
    "Correctly clears 63% of safe buildings — reduces unnecessary inspections",
    "If model says safe, it is correct 99% of the time — key for prioritization",
    "Tuned to prioritize recall over precision given the cost of missed failures"
  )
) %>%
  kable(align = "lll") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), font_size = 19) %>%
  row_spec(4, bold = TRUE, background = "#e8f5e9")
```

------------------------------------------------------------------------

## Where Does the Model Struggle? Residual Map

```{r residual-map}
#| fig-align: center
#| out-width: 85%

insp_coords <- inspections_fire_opa %>%
  st_transform(crs = 4326) %>%
  mutate(
    lon = st_coordinates(.)[, 1],
    lat = st_coordinates(.)[, 2]
  ) %>%
  st_drop_geometry()

test_data_sf <- test_data %>%
  left_join(insp_coords %>% select(objectd, lon, lat), by = "objectd") %>%
  mutate(residual = fail - predicted_prob) %>%
  filter(!is.na(residual), !is.na(lon), !is.na(lat)) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

ggplot() +
  geom_sf(data = block_groups, fill = "grey90", color = "white", linewidth = 0.1) +
  geom_sf(data = test_data_sf, aes(color = residual), size = 0.5, alpha = 0.6) +
  scale_color_gradient2(
    low = "#2166ac", mid = "white", high = "#d6604d",
    midpoint = 0, name = "Residual\n(actual − predicted)"
  ) +
  labs(
    title    = "Spatial Distribution of Model Residuals",
    subtitle = "Red = under-predicted risk · Blue = over-predicted risk · Moran's I confirms residual spatial autocorrelation",
    caption  = "Source: OPA Philadelphia"
  ) +
  mapTheme
```

------------------------------------------------------------------------

## Limitations and Bias {.scrollable .smaller}

**The training data reflects enforcement, not true risk**\
The model learns from where the city *already* inspects—complaint-driven neighborhoods are overrepresented. It captures enforcement patterns, not underlying building conditions.

**Underreporting compounds over time**\
Vulnerable communities call 311 less → lower predicted risk → fewer inspections → continued under-enforcement. The `Pct_Renter_Occupied × log_n_requests` interaction captures this partially but does not resolve it.

**Certification type dominates (OR = 3.66)**\
Sprinkler inspections fail at 3.7× the baseline rate. This is a legitimate control variable but means the model is partly predicting *what kind of inspection was scheduled*, not just building condition.

**Residual spatial autocorrelation**\
Moran's I on residuals shows remaining spatial structure, suggesting spatial cross-validation and additional spatial features as clear next steps.

> **This is a decision-support tool. It should inform inspector deployment, not replace human judgment.**

------------------------------------------------------------------------

## Implementation and Safeguards {.scrollable .smaller}

**How the city uses this:**

1.  Run final model on all L&I-registered properties periodically
2.  Assign predicted failure probability (0–1) to each building
3.  Sort into risk tiers (High / Medium / Low) by district
4.  Produce prioritized inspection lists for field staff
5.  Retrain quarterly as new inspection outcomes accumulate

**Required safeguards:**

-   Human oversight: scores inform, not replace, inspector decisions
-   Equity audits by census tract: monitor for systematic over/under-prediction in lower-income areas
-   Transparent methodology: public documentation and grievance process
-   Model monitoring: flag degradation in cross-validated AUC over time

------------------------------------------------------------------------

## References {.smaller}

**Data Sources**

-   Philadelphia Department of Licenses & Inspections - Building Certifications (OpenDataPhilly)
-   City of Philadelphia Office of Property Assessment - `opa_properties_public.geojson`
-   Philadelphia 311 Service Requests - Fire Department calls (OpenDataPhilly)
-   U.S. Census Bureau - American Community Survey 5-Year Estimates, 2022, block group level

**References**

-   Pew Charitable Trusts (2021). *Philadelphia's Rental Housing Challenge.*
-   Urban Institute (2022). *Who Is Left Behind by Complaint-Based Code Enforcement?*
-   National Low Income Housing Coalition (2022). *Out of Reach: The High Cost of Housing.*

**R Packages**

`tidyverse` · `sf` · `tidycensus` · `tigris` · `caret` · `pROC` · `spdep` · `kableExtra`
