---
title: "Lab 4: Spatial Predictive Analysis"
author: "Arzoo Gupta"
format: html
editor: visual
---

\***AI (Claude Sonnet 4.6 was used to debug code)**

## **A note on predictive policing:**

This is an academic exercise conducted to understand the mechanics, and the limits, of spatial predictive modeling in a public safety context. Building a model that predicts crime risk is not the same as recommending that one be used, and this lab should be read in that spirit.

Chicago offers one of the most well-documented cautionary examples of what happens when predictive tools move from the classroom into deployment without sufficient scrutiny. It was a story about bad data — and the consent decree is where that becomes legally verifiable. The DOJ investigation that preceded the consent decree was triggered by the 2014 killing of Laquan McDonald. That investigation found that the Chicago Police Department routinely violated the constitutional rights of Black and Latinx residents, used excessive force, and operated with a culture of impunity — and that officers commonly stopped and searched Black youth without legal justification, driven by racial stereotypes rather than reasonable suspicion. [[The Chicago Reporter]{.underline}](https://www.chicagoreporter.com/six-years-23-percent-chicagos-failed-promise-of-police-reform/) The Illinois Attorney General filed a lawsuit in August 2017 to force a consent decree, and it took effect in January 2019. [[Chicagopoliceconsentdecree]{.underline}](https://www.chicagopoliceconsentdecree.org/about/)

This is where a landmark 2019 report by the AI Now Institute becomes directly relevant. The researchers examined 13 jurisdictions that had used predictive policing systems and cross-referenced them with consent decrees and federal court-monitored settlements — using those legal findings as proof that unconstitutional policing practices had shaped the underlying data. Chicago was their first case study: a documented example of a jurisdiction where dirty data was ingested directly into the city's predictive policing system. [[NYULawReview]{.underline}](https://www.nyulawreview.org/wp-content/uploads/2019/04/NYULawReview-94-Richardson-Schultz-Crawford.pdf) The concept they introduced — "dirty data" — refers to data derived from or influenced by corrupt, biased, and unlawful practices, including both discriminatory policing and manipulation of crime statistics. [[Brennan Center for Justice]{.underline}](https://www.brennancenter.org/our-work/research-reports/predictive-policing-explained) The consent decree provided the legally verified paper trail proving that this is exactly what happened in Chicago.

This matters for the analysis that follows. The burglary data used in this lab was generated by a police department that was simultaneously under federal scrutiny for unconstitutional practices. Any model trained on that data does not neutrally describe where crime occurs — it describes where a constitutionally non-compliant police department directed its attention and made its arrests.

After six years of federal oversight and more than \$15 million spent on independent monitoring, CPD has fully implemented just 23 percent of the reforms it agreed to. [[The Chicago Reporter]{.underline}](https://www.chicagoreporter.com/six-years-23-percent-chicagos-failed-promise-of-police-reform/) The data being generated today is still produced by an institution mid-reform at best. What follows is an exercise in understanding how spatial predictive models are constructed — offered precisely so we can think critically about why they should never be deployed uncritically, and why Chicago's experience is not an anomaly but a warning.

## Introduction

When alley lights go out, they don't just create inconvenience; they create voids of darkness in spaces that already lack foot traffic and natural surveillance. This analysis examines 311 reports of broken alley lights in Chicago during 2017 to ask a spatial question: can the geographic distribution of infrastructure neglect — specifically, unlit alleys — help predict where burglaries occur?

But before proceeding, it is worth naming what this kind of analysis does and does not do. Mapping broken alley lights and burglaries together does not reveal causation — it reveals correlation.

Using 311 data as a predictor of crime risk means using a signal that is itself shaped by who calls 311, how quickly complaints are resolved, and which neighborhoods receive timely responses. A model trained on this data doesn't neutrally describe reality. That is the data problem at the heart of any predictive policing application, and it is why we build this model to understand.

The neighborhoods with the most broken light complaints are also the neighborhoods that could have:

1.  Historically been under-resourced, over-policed, and subject to discriminatory city services.
2.  Have an active community that has found the 311 service reliable.

```{r setup}
#| message: false
#| warning: false
#| echo: false


# Load required packages
library(tidyverse)      # Data manipulation
library(sf)             # Spatial operations
library(here)           # Relative file paths
library(viridis)        # Color scales
library(terra)          # Raster operations (replaces 'raster')
library(spdep)          # Spatial dependence
library(FNN)            # Fast nearest neighbors
library(MASS)           # Negative binomial regression
library(patchwork)      # Plot composition (replaces grid/gridExtra)
library(knitr)          # Tables
library(kableExtra)     # Table formatting
library(classInt)       # Classification intervals
library(lubridate)


# Spatstat for analyzing point patterns - split into sub-packages
library(spatstat.geom)    # Spatial geometries
library(spatstat.explore) # Spatial exploration/KDE

# Set options
options(scipen = 999)  # No scientific notation
set.seed(5080)         # Reproducibility, could be any number

# Create consistent theme for visualizations
theme_crime <- function(base_size = 11) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", size = base_size + 1),
      plot.subtitle = element_text(color = "gray30", size = base_size - 1),
      legend.position = "right",
      panel.grid.minor = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

# Set as default
theme_set(theme_crime())

cat("✓ All packages loaded successfully!\n")
cat("✓ Working directory:", getwd(), "\n")

```

### Required Components

-   [ ] **Introduction**: Brief description of your chosen 311 violation type and why you selected it

-   [x] **Data Exploration**: Maps and summary statistics of your violation data

-   [x] **Fishnet Grid**: Successfully created and visualized grid with aggregated counts

-   [x] **Spatial Features**: At least 2-3 spatial features calculated (k-NN, distance measures, etc.)

-   [x] **Local Moran’s I**: Spatial autocorrelation analysis with interpretation

-   [x] **Count Regression**: Both Poisson and Negative Binomial models with comparison

-   [x] **Spatial Cross-Validation**: LOGO CV implemented on 2017 data with reported metrics

-   [ ] **Temporal Validation**: 2018 data aggregated and tested with 2017 model(OPTIONAL.)

-   [ ] **Validation Comparison**: Comparison of spatial vs. temporal validation performance

-   [ ] **Model Comparison**: Your model vs. KDE baseline for 2018 predictions

-   [ ] **Error Analysis**: Maps and discussion of where model performs well/poorl

-   [ ] **Written Explanations**: Your own words explaining each step

## Exploratory Data Analysis for Broken Alley Lights

### Part 1: Data Loading & Exploration

The 311 violation data for broken alley lights have been loaded into the R environment. Along with this Police Beats, Districts, and the Chicago Boundary have been using API's from the city of Chicago's data portal. The data has been subsequently cleaned for further analysis. It has also been projected into the Chicago 1983 NAD state-plane to ensure spatial accuracy in future calculations.

```{r cleaning-violation-data}
#| message: false
#| warning: false
#| echo: false



req_311 <- read.csv("Data/311_req_alley_lights.csv")

#cleaning the data to conduct EDA

#this has all the alley lights complaints from 2010, I'm getting a dataset with only complaints in 2017 

lights_out_2017 <- req_311%>%
  mutate(date = mdy(Creation.Date)) %>%          
  filter(year(date) == 2017) %>%         
  mutate(date = format(date, "%Y/%m/%d"))

#adding cooridnate system 
lights_out_2017 <- lights_out_2017 %>%
  filter(!is.na(Longitude) & !is.na(Latitude)) %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  st_transform("ESRI:102271")

# Load police districts (used for spatial cross-validation)
policeDistricts <- 
  st_read("https://data.cityofchicago.org/api/geospatial/24zt-jpfn?method=export&format=GeoJSON", quiet = T) %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(District = dist_num)

# Load police beats (smaller administrative units)

policeBeats <- 
  st_read("https://data.cityofchicago.org/api/geospatial/n9it-hstw?method=export&format=GeoJSON", quiet = T) %>%
  st_transform('ESRI:102271') %>%
  dplyr::select(Beat = beat_num)

# Load Chicago boundary
chicagoBoundary <- 
  st_read("https://raw.githubusercontent.com/urbanSpatial/Public-Policy-Analytics-Landing/master/DATA/Chapter5/chicagoBoundary.geojson", quiet = T) %>%
  st_transform('ESRI:102271')

cat("✓ Cleaned Alley Lights Violation Data\n")
cat("  - Violations:", nrow(lights_out_2017), "\n")
cat("  - Date range:", format(min(lights_out_2017$date, na.rm = TRUE)), "to", 
    format(max(lights_out_2017$date, na.rm = TRUE)), "\n")
cat("✓ Loaded spatial boundaries\n")
cat("  - Police districts:", nrow(policeDistricts), "\n")
cat("  - Police beats:", nrow(policeBeats), "\n")

```

#### Choropleth Map of all non-functioning alley lights 311 calls

The dot map below reveals something more important than the total count: these complaints are not evenly distributed. Large swaths of the North Side and the lakefront appear relatively sparse, while the West Side and parts of the South Side are so densely covered in dots that individual points are nearly indistinguishable. This is not a map of broken lights, it is a map of where broken lights are reported, where residents engage with city services.

```{r}
#| message: false
#| warning: false
#| echo: false

#creating a choropleth map of all violations in the year 2017
ggplot() +
  geom_sf(data = chicagoBoundary, fill = "gold", color = "white") +
  geom_sf(data = lights_out_2017, size = 0.3, alpha = 0.4, color = "black") +
  labs(title = "Alley Light Violations, Chicago 2017",
       subtitle = paste("Total Violations:", nrow(lights_out_2017)))+
  theme(panel.grid.major = element_blank())

```

#### Bar Charts for 311 calls

The monthly bar chart complicates this further. Rather than peaking in summer, complaints are actually highest in January, February, and March—the winter months—and notably lower through July and August. This is counterintuitive if we think of 311 calls as a direct response to dangerous darkness. It may instead reflect patterns of indoor activity in winter leading people to notice alley lights from windows, or seasonal city maintenance cycles. Whatever the cause, the seasonality is a reminder that 311 data is a record of human behavior around reporting, not a direct sensor reading of infrastructure failure.

The resolution time histogram drives this point home sharply. The distribution is heavily right-skewed: the vast majority of complaints are resolved within a few days, creating the tall spike at the left of the histogram, but a long tail extends far to the right—some complaints take months to close. This variance in response time is not random. It is spatially structured, as the district-level maps below reveal.

```{r EDA-Violation-Data}
#| message: false
#| warning: false
#| echo: false 


#creating a bar chart of all violations over the 12 months 

lights_out_2017 %>%
  st_drop_geometry() %>%
  mutate(month = month(ymd(date), label = TRUE)) %>%
  group_by(month) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = month, y = count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Alley Light Violations by Month",
       subtitle = "Chicago 2017",
       x = "Month", y = "Number of Violations")


lights_out_2017 <- lights_out_2017 %>%
  mutate(resolution_days = as.numeric(mdy(Completion.Date) - mdy(Creation.Date)))

lights_out_2017 %>%
  st_drop_geometry() %>%
  ggplot(aes(x = resolution_days)) +
  geom_histogram(binwidth = 5, fill = "yellow", color = "black") +
  labs(title = "Distribution of Resolution Times",
       subtitle = "Chicago Alley Light Violations 2017",
       x = "Days to Resolve", y = "Count")

```

#### Choropleth Maps of Resolution Time and Multiple Complaints

The left district map shows average resolution time by police district. The differences are substantial with some districts resolve complaints in under three weeks on average while others stretch past 60 days.

The right map shows the duplicate complaint rate: the share of 311 calls flagged as duplicates, meaning another resident already reported the same broken light. Districts with the highest duplicate rates, reaching 40–50% in some areas, are places where lights stay broken long enough for multiple residents to independently call it in. Together these two maps document something the aggregate count hides: the city's response to broken alley lights is deeply unequal, and that inequality follows the same geography as Chicago's broader patterns of disinvestment. The districts that wait longest for repairs and generate the most duplicate complaints are the same neighborhoods that will show up later in this analysis as burglary hot spots. That is not a coincidence, it is a structural condition that any model trained on this data will absorb and reproduce.

```{r comparing-violations}
#| message: false
#| warning: false
#| echo: false 


resolution_time <-lights_out_2017 %>%
  st_join(policeDistricts) %>%
  st_drop_geometry() %>%
  group_by(District) %>%
  summarise(avg_resolution = mean(resolution_days, na.rm = TRUE)) %>%
  right_join(policeDistricts, by = "District") %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = avg_resolution), color = "white") +
  scale_fill_viridis(name = "Avg Days\nto Resolve") +
  labs(subtitle = "Average Alley Light Resolution Time") +
  theme(panel.grid.major = element_blank())


mult_complaints <- lights_out_2017 %>%
  st_join(policeDistricts) %>%
  st_drop_geometry() %>%
  mutate(is_duplicate = str_detect(Status, "Dup")) %>%
  group_by(District) %>%
  summarise(dup_rate = mean(is_duplicate, na.rm = TRUE)) %>%
  right_join(policeDistricts, by = "District") %>%
  st_as_sf() %>%
  ggplot() +
  geom_sf(aes(fill = dup_rate), color = "white") +
  scale_fill_viridis(name = "Duplicate\nRate", labels = scales::percent) +
  labs(subtitle = "Multiple Violation Reporting Rate",
       ,theme = theme(plot.title = element_text(face = "italic"))) +
  theme(panel.grid.major = element_blank())

resolution_time + mult_complaints +
  plot_annotation(
    title = "Alley Light Violations by Police District, Chicago 2017", 
    theme = theme(plot.title = element_text(face = "bold")))

```

### Part 2: Fishnet Grid Creation

Rather than analyzing violations as individual points, we aggregate them into a regular grid of 500-meter by 500-meter cells covering the entire city. This fishnet approach converts Chicago into a set of consistently-sized units of analysis compatible with regression modeling. Instead of asking "how many lights are broken near this address," we can ask "how many lights are broken in each cell" a format that allows us to attach predictors and outcomes to the same spatial unit and estimate a statistical relationship between them.

```{r fishnet}
#| message: false
#| warning: false
#| echo: false


# Create 500m x 500m grid
fishnet <- st_make_grid(
  chicagoBoundary,
  cellsize = 500,  # 500 meters per cell
  square = TRUE
) %>%
  st_sf() %>%
  mutate(uniqueID = row_number()) %>%
  st_transform('ESRI:102271') 

# Keep only cells that intersect Chicago
fishnet <- fishnet[chicagoBoundary, ]

# View basic info
cat("✓ Created fishnet grid\n")
cat("  - Number of cells:", nrow(fishnet), "\n")
cat("  - Cell size:", 500, "x", 500, "meters\n")
cat("  - Cell area:", round(st_area(fishnet[1,])), "square meters\n")

```

```{r fishnet-alley-lights}
#| message: false
#| warning: false
#| echo: false

# Aggregate alley light violations to fishnet
lights_out_fishnet <- st_join(lights_out_2017, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(lights_out_2017 = n())

# Join to fishnet
fishnet <- fishnet %>%
  left_join(lights_out_fishnet, by = "uniqueID") %>%
  mutate(lights_out_2017 = replace_na(lights_out_2017, 0))

cat("Broken Light distribution:\n")
summary(fishnet$lights_out_2017)

```

The fishnet map of broken alley light counts reveals a pattern that the dense dot map above obscured. Most of the city's cells contain relatively few complaints—the map is predominantly dark purple—but a handful of cells, particularly in clusters on the West Side, register counts approaching or exceeding 200. These extreme cells are not randomly distributed; they sit in recognizable corridors of the city that correspond to areas with the longest resolution times and highest duplicate complaint rates we saw in the district maps. The fishnet makes this structure legible for modeling in a way the point map could not.

t also makes a quiet data quality problem visible. The North Side and lakefront cells are almost uniformly dark, near-zero complaint counts, while structurally similar streets in other parts of the city register far more. This could reflect genuinely better infrastructure maintenance. Or it could reflect lower reporting rates, faster city response that resolves complaints before they accumulate, or simply less trust in 311 as a useful channel.

```{r fishnet-violation-viz}
#| message: false
#| warning: false
#| echo: false

ggplot() +
  geom_sf(data = fishnet, aes(fill = lights_out_2017), color = NA) +
  scale_fill_viridis_c(name = "Count", option = "inferno") +
  labs(title = "Broken Alley Lights 311 Calls") +
  theme_crime()

```

### Part 3: Spatial Features

#### K-Nearest Neighbours

A single count of broken lights in each cell captures local conditions but misses the spatial context surrounding that cell. A grid cell with zero complaints is not necessarily a well-lit area, it may simply be a place where the reporting behavior or response times we documented above result in fewer recorded complaints. To capture proximity to known infrastructure failures regardless of what happens to be recorded in any individual cell, we calculate the mean distance from each cell's centroid to its three nearest broken light complaints. A small value means the cell sits close to documented failures; a large value means it is relatively isolated from them. This feature allows the model to borrow spatial information from neighboring areas and is particularly important for smoothing over the reporting-behavior noise that makes the raw counts unreliable in some parts of the city.

```{r nearest-neighbors}
#| warning: false
#| echo: false

# Calculate mean distance to 3 nearest broken lights
# (Do this OUTSIDE of mutate to avoid sf conflicts)

# Get coordinates
fishnet_coords <- st_coordinates(st_centroid(fishnet))
lights_out_coords <- st_coordinates(lights_out_2017)

# Calculate k nearest neighbors and distances
nn_result <- get.knnx(lights_out_coords, fishnet_coords, k = 3)

# Add to fishnet
fishnet <- fishnet %>%
  mutate(
    lights_out.nn = rowMeans(nn_result$nn.dist)
  )

cat("✓ Calculated nearest neighbor distances\n")
summary(fishnet$lights_out.nn)


# Get centroids of "High-High" cells (hot spots)

```

#### Local Moran's I

Local Moran's I is a measure of spatial autocorrelation. It asks whether a cell's value is similar to or different from its neighbors in a way that is statistically unlikely to occur by chance. Applied here to the broken alley lights count, it classifies each cell into a cluster type.

```{r local-moransI}
#| message: false
#| warning: false
#| echo: false

# Function to calculate Local Moran's I
calculate_local_morans <- function(data, variable, k = 5) {
  
  # Create spatial weights
  coords <- st_coordinates(st_centroid(data))
  neighbors <- knn2nb(knearneigh(coords, k = k))
  weights <- nb2listw(neighbors, style = "W", zero.policy = TRUE)
  
  # Calculate Local Moran's I
  local_moran <- localmoran(data[[variable]], weights)
  
  # Classify clusters
  mean_val <- mean(data[[variable]], na.rm = TRUE)
  
  data %>%
    mutate(
      local_i = local_moran[, 1],
      p_value = local_moran[, 5],
      is_significant = p_value < 0.05,
      
      moran_class = case_when(
        !is_significant ~ "Not Significant",
        local_i > 0 & .data[[variable]] > mean_val ~ "High-High",
        local_i > 0 & .data[[variable]] <= mean_val ~ "Low-Low",
        local_i < 0 & .data[[variable]] > mean_val ~ "High-Low",
        local_i < 0 & .data[[variable]] <= mean_val ~ "Low-High",
        TRUE ~ "Not Significant"
      )
    )
}

# Apply to alley light
fishnet <- calculate_local_morans(fishnet, "lights_out_2017", k = 5)
```

The resulting map is interesting. Rather than a broad wash of hot spots across the South and West sides, the High-High cells, places where many broken lights are surrounded by more broken lights, appear as a series of tight, discrete clusters scattered along a north-south corridor through the middle of the city, with additional clusters on the West Side. The vast majority of the city, including most of the North Side and far South Side, is classified as not significant.

This spottiness is itself informative. It tells us that infrastructure neglect as captured by 311 data is not a smooth citywide gradient but a pattern of localized clusters, specific corridors and intersections where failures accumulate and persist. It also tells us that the relationship between broken lights and the underlying conditions we are trying to capture is more spatially concentrated than a simple choropleth of complaint counts would suggest. These hot spots will become a key predictive feature: whether a cell sits near one of these clusters of persistent failure is a meaningful signal about the environmental conditions in that area.

```{r morans-viz}
#| fig-width: 8
#| fig-height: 6
#| echo: false

# Visualize hot spots
ggplot() +
  geom_sf(
    data = fishnet, 
    aes(fill = moran_class), 
    color = NA
  ) +
  scale_fill_manual(
    values = c(
      "High-High" = "#d7191c",
      "High-Low" = "#fdae61",
      "Low-High" = "#abd9e9",
      "Low-Low" = "#2c7bb6",
      "Not Significant" = "gray90"
    ),
    name = "Cluster Type"
  ) +
  labs(
    title = "Local Moran's I: Broken Alley Lights Cluster",
    subtitle = "High-High = Hot spots of disorder"
  ) +
  theme_crime()

```

#### Distance to Hotspots

Using the High-High clusters identified above, we calculate the Euclidean distance from each grid cell to the nearest broken lights hot spot. This converts the cluster map into a continuous feature: cells within or immediately adjacent to a hot spot get near-zero values, while cells far from any significant cluster get large values. It is a concise summary of how embedded each cell is in a zone of systemic infrastructure neglect, and it is the spatial feature most likely to carry predictive power across the full fishnet even in cells that have few direct complaints of their own.

```{r distance-to-hotspots}
#| warning: false
#| echo: false

hotspots <- fishnet %>%
  filter(moran_class == "High-High") %>%
  st_centroid()

# Calculate distance from each cell to nearest hot spot
if (nrow(hotspots) > 0) {
  fishnet <- fishnet %>%
    mutate(
      dist_to_hotspot = as.numeric(
        st_distance(st_centroid(fishnet), hotspots %>% st_union())
      )
    )
   cat("✓ Calculated distance to broken alley lights hot spots\n")
  cat("  - Number of hot spot cells:", nrow(hotspots), "\n")
} else {
  fishnet <- fishnet %>%
    mutate(dist_to_hotspot = 0)
  cat("⚠ No significant hot spots found\n")
}


```

### Part 4: Loading Burglary Data

```{r burglary-data}
#| message: false
#| echo: false

# Load from provided data file (downloaded from Chicago open data portal)
burglaries <- st_read("Data/burglaries.shp", quiet = T) %>% 
  st_transform('ESRI:102271')


cat("\n✓ Loaded burglary data\n")
cat("  - Number of burglaries:", nrow(burglaries), "\n")
cat("  - CRS:", st_crs(burglaries)$input, "\n")
cat("  - Date range:", format(min(burglaries$Date, na.rm = TRUE)), "to", 
    format(max(burglaries$Date, na.rm = TRUE)), "\n")

p1 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_sf(data = burglaries, color = "#d62828", size = 0.1, alpha = 0.4) +
  labs(
    title = "Burglary Locations",
    subtitle = paste0("Chicago 2017, n = ", nrow(burglaries))
  )

p2 <- ggplot() + 
  geom_sf(data = chicagoBoundary, fill = "gray95", color = "gray60") +
  geom_density_2d_filled(
    data = data.frame(st_coordinates(burglaries)),
    aes(X, Y),
    alpha = 0.7,
    bins = 8
  ) +
  scale_fill_viridis_d(
    option = "plasma",
    direction = -1,
    guide = "none"  # Modern ggplot2 syntax (not guide = FALSE)
  ) +
  labs(
    title = "Density Surface",
    subtitle = "Kernel density estimation"
  )


# Combine plots using patchwork (modern approach)
p1 + p2 + 
  plot_annotation(
    title = "Spatial Distribution of Burglaries in Chicago",
    tag_levels = 'A'
  )
```

```{r comparing-burglary}
#| fig-width: 8
#| fig-height: 6
#| echo: false
#| warning: false
#| message: false

# Spatial join: which cell contains each burglary?
burglaries_fishnet <- st_join(burglaries, fishnet, join = st_within) %>%
  st_drop_geometry() %>%
  group_by(uniqueID) %>%
  summarize(countBurglaries = n()) 


st_crs(burglaries_fishnet)

# Join back to fishnet (cells with 0 burglaries will be NA)
fishnet <- fishnet %>%
  left_join(burglaries_fishnet, by = "uniqueID") %>%
  mutate(countBurglaries = replace_na(countBurglaries, 0))

# Summary statistics
cat("\nBurglary count distribution:\n")
summary(fishnet$countBurglaries)
cat("\nCells with zero burglaries:", 
    sum(fishnet$countBurglaries == 0), 
    "/", nrow(fishnet),
    "(", round(100 * sum(fishnet$countBurglaries == 0) / nrow(fishnet), 1), "%)\n")

ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "Burglaries",
    option = "plasma",
    trans = "sqrt",  # Square root for better visualization of skewed data
    breaks = c(0, 1, 5, 10, 20, 40)
  ) +
  labs(
    title = "Burglary Counts by Grid Cell",
    subtitle = "500m x 500m cells, Chicago 2017"
  ) +
  theme_crime()
```

### Part 5: KDE Baseline

```{r kde-baseline}
#| message: false
#| warning: false
#| echo: false

# Convert burglaries to ppp (point pattern) format for spatstat
burglaries_ppp <- as.ppp(
  st_coordinates(burglaries),
  W = as.owin(st_bbox(chicagoBoundary))
)

# Calculate KDE with 1km bandwidth
kde_burglaries <- density.ppp(
  burglaries_ppp,
  sigma = 1000,  # 1km bandwidth
  edge = TRUE    # Edge correction
)

# Convert to terra raster (modern approach, not raster::raster)
kde_raster <- rast(kde_burglaries)

# Extract KDE values to fishnet cells
fishnet <- fishnet %>%
  mutate(
    kde_value = terra::extract(
      kde_raster,
      vect(fishnet),
      fun = mean,
      na.rm = TRUE
    )[, 2]  # Extract just the values column
  )

cat("✓ Calculated KDE baseline\n")

```

```{r kde-viz}
#| fig-width: 8
#| fig-height: 6
#| echo: false

kde <-ggplot() +
  geom_sf(data = fishnet, aes(fill = kde_value), color = NA) +
  geom_sf(data = chicagoBoundary, fill = NA, color = "white", linewidth = 1) +
  scale_fill_viridis_c(
    name = "KDE Value",
    option = "plasma"
  ) +
  labs(
    title = "Kernel Density Estimation Baseline",
    subtitle = "Simple spatial smoothing of burglary locations"
  ) +
  theme_crime()

p1 + kde + 
  plot_annotation(
    title = "Spatial Distribution of Burglaries in Chicago",
    tag_levels = 'A'
  )
```

```{r comparing-violation-burglary}
#| fig-width: 10
#| fig-height: 4
#| echo: false

p1 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = lights_out_2017), color = NA) +
  scale_fill_viridis_c(name = "Count", option = "plasma") +
  labs(title = "Broken Alley Light 311 Calls") +
  theme_crime()

p2 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  scale_fill_viridis_c(name = "Count", option = "plasma") +
  labs(title = "Burglaries") +
  theme_crime()

p1 + p2 +
  plot_annotation(title = "Are broken lights and burglaries correlated?")

```

```{r join-police-dsitricts}
#| echo: false
#| warning: false
#| message: false

# Join district information to fishnet
fishnet <- st_join(
  fishnet,
  policeDistricts,
  join = st_within,
  left = TRUE
) %>%
  filter(!is.na(District))  # Remove cells outside districts

cat("✓ Joined police districts\n")
cat("  - Districts:", length(unique(fishnet$District)), "\n")
cat("  - Cells:", nrow(fishnet), "\n")

```

### Part 6: Count Regression Models

```{r prepare-data}
#| echo: false
#| warning: false
#| message: false

# Create clean modeling dataset
fishnet_model <- fishnet %>%
  st_drop_geometry() %>%
  dplyr::select(
    uniqueID,
    District,
    countBurglaries,
   lights_out_2017,
    lights_out.nn,
    dist_to_hotspot
  ) %>%
  na.omit()  # Remove any remaining NAs

cat("✓ Prepared modeling data\n")
cat("  - Observations:", nrow(fishnet_model), "\n")
cat("  - Variables:", ncol(fishnet_model), "\n")

```

```{r poisson-regression}
#| echo: false
#| warning: false
#| message: false

# Fit Poisson regression
model_poisson <- glm(
  countBurglaries ~ lights_out_2017 + lights_out.nn + 
    dist_to_hotspot,
  data = fishnet_model,
  family = "poisson"
)

# Summary
summary(model_poisson)

```

Checking for over disperson

```{r dispersion}
#| echo: false
#| warning: false
#| message: false


dispersion <- sum(residuals(model_poisson, type = "pearson")^2) / 
              model_poisson$df.residual

cat("Dispersion parameter:", round(dispersion, 2), "\n")
cat("Rule of thumb: >1.5 suggests overdispersion\n")

if (dispersion > 1.5) {
  cat("⚠ Overdispersion detected! Consider Negative Binomial model.\n")
} else {
  cat("✓ Dispersion looks okay for Poisson model.\n")
}

```

Fitting Negative Binomial Model

```{r negative-binomial-model}
#| echo: false
#| warning: false
#| message: false


# Fit Negative Binomial model
model_nb <- glm.nb(
  countBurglaries ~ lights_out_2017 + lights_out.nn + 
    dist_to_hotspot,
  data = fishnet_model
)

# Summary
summary(model_nb)

# Compare AIC (lower is better)
cat("\nModel Comparison:\n")
cat("Poisson AIC:", round(AIC(model_poisson), 1), "\n")
cat("Negative Binomial AIC:", round(AIC(model_nb), 1), "\n")

```

### Part 7: Spatial Cross-Validation (2017)

```{r LOGO}
#| warning: false
#| message: false
#| echo: false

districts <- unique(fishnet_model$District)
cv_results <- tibble()

cat("Running LOGO Cross-Validation...\n")

for (i in seq_along(districts)) {
  
  test_district <- districts[i]
  
  # Split data
  train_data <- fishnet_model %>% filter(District != test_district)
  test_data <- fishnet_model %>% filter(District == test_district)
  
  # Fit model on training data
  model_cv <- glm.nb(
    countBurglaries ~ lights_out_2017+ lights_out.nn + 
      dist_to_hotspot,
    data = train_data
  )
  
  # Predict on test data
  test_data <- test_data %>%
    mutate(
      prediction = predict(model_cv, test_data, type = "response")
    )
  
  # Calculate metrics
  mae <- mean(abs(test_data$countBurglaries - test_data$prediction))
  rmse <- sqrt(mean((test_data$countBurglaries - test_data$prediction)^2))
  
  # Store results
  cv_results <- bind_rows(
    cv_results,
    tibble(
      fold = i,
      test_district = test_district,
      n_test = nrow(test_data),
      mae = mae,
      rmse = rmse
    )
  )
  
  cat("  Fold", i, "/", length(districts), "- District", test_district, 
      "- MAE:", round(mae, 2), "\n")
}

# Overall results
cat("\n✓ Cross-Validation Complete\n")
cat("Mean MAE:", round(mean(cv_results$mae), 2), "\n")
cat("Mean RMSE:", round(mean(cv_results$rmse), 2), "\n")

```

```{r cv-results}
#| echo: false
#| warning: false
#| message: false

# Show results
cv_results %>%
  arrange(desc(mae)) %>%
  kable(
    digits = 2,
    caption = "LOGO CV Results by District"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))


```

### Part 8: Model Evaluation

```{r}
#| echo: false
#| warning: false
#| message: false


final_model <- glm.nb(
  countBurglaries ~ lights_out_2017 + lights_out.nn + 
    dist_to_hotspot,
  data = fishnet_model
)

# Add predictions back to fishnet
fishnet <- fishnet %>%
  mutate(
    prediction_nb = predict(final_model, fishnet_model, type = "response")[match(uniqueID, fishnet_model$uniqueID)]
  )

# Also add KDE predictions (normalize to same scale as counts)
kde_sum <- sum(fishnet$kde_value, na.rm = TRUE)
count_sum <- sum(fishnet$countBurglaries, na.rm = TRUE)
fishnet <- fishnet %>%
  mutate(
    prediction_kde = (kde_value / kde_sum) * count_sum
  )



```

```{r compare-models -viz}
#| fig-width: 12
#| fig-height: 4
#| echo: false
#| warning: false
#| message: false


# Create three maps
p1 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = countBurglaries), color = NA) +
  scale_fill_viridis_c(name = "Count", option = "plasma", limits = c(0, 15)) +
  labs(title = "Actual Burglaries") +
  theme_crime()

p2 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_nb), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "Model Predictions (Neg. Binomial)") +
  theme_crime()

p3 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = prediction_kde), color = NA) +
  scale_fill_viridis_c(name = "Predicted", option = "plasma", limits = c(0, 15)) +
  labs(title = "KDE Baseline Predictions") +
  theme_crime()

p1 + p2 + p3 +
  plot_annotation(
    title = "Actual vs. Predicted Burglaries",
    subtitle = "Does our complex model outperform simple KDE?"
  )
```

```{r comaring-model-metrics}
#| echo: false
#| warning: false
#| message: false


# Calculate performance metrics
comparison <- fishnet %>%
  st_drop_geometry() %>%
  filter(!is.na(prediction_nb), !is.na(prediction_kde)) %>%
  summarize(
    model_mae = mean(abs(countBurglaries - prediction_nb)),
    model_rmse = sqrt(mean((countBurglaries - prediction_nb)^2)),
    kde_mae = mean(abs(countBurglaries - prediction_kde)),
    kde_rmse = sqrt(mean((countBurglaries - prediction_kde)^2))
  )

comparison %>%
  pivot_longer(everything(), names_to = "metric", values_to = "value") %>%
  separate(metric, into = c("approach", "metric"), sep = "_") %>%
  pivot_wider(names_from = metric, values_from = value) %>%
  kable(
    digits = 2,
    caption = "Model Performance Comparison"
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover"))

```

### Part 9: Model Errors

```{r}
#| fig-width: 10
#| fig-height: 5
#| echo: false
#| warning: false
#| message: false


# Calculate errors
fishnet <- fishnet %>%
  mutate(
    error_nb = countBurglaries - prediction_nb,
    error_kde = countBurglaries - prediction_kde,
    abs_error_nb = abs(error_nb),
    abs_error_kde = abs(error_kde)
  )

# Map errors
p1 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = error_nb), color = NA) +
  scale_fill_gradient2(
    name = "Error",
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0,
    limits = c(-10, 10)
  ) +
  labs(title = "Model Errors (Actual - Predicted)") +
  theme_crime()

p2 <- ggplot() +
  geom_sf(data = fishnet, aes(fill = abs_error_nb), color = NA) +
  scale_fill_viridis_c(name = "Abs. Error", option = "plasma") +
  labs(title = "Absolute Model Errors") +
  theme_crime()

p1 + p2
```

```{r}
#| echo: false
#| warning: false
#| message: false

# Create nice summary table
model_summary <- broom::tidy(final_model, exponentiate = TRUE) %>%
  mutate(
    across(where(is.numeric), ~round(., 3))
  )

model_summary %>%
  kable(
    caption = "Final Negative Binomial Model Coefficients (Exponentiated)",
    col.names = c("Variable", "Rate Ratio", "Std. Error", "Z", "P-Value")
  ) %>%
  kable_styling(bootstrap_options = c("striped", "hover")) %>%
  footnote(
    general = "Rate ratios > 1 indicate positive association with burglary counts."
  )
```
